Modelling state-switching functional data with hidden Markov models: reviewer 2 response

### 2021-07-09 Report/Sections/intro.tex
Final edits

FDA was originally developed to process curves assumed to be independent replicates (i.e., there is no between-curve dependence), and within-curve fine-scale structure is not usually incorporated in FDA models. However, sets of curves often exhibit complex sequential dependencies both between curves and within curves, especially in the case of biologging data \citep{Barajas:2017}.
On a coarse scale, dive profiles of marine animals show a discrete number of distinct dive ``types"
%that aggregate together in time 
where the sequence of dive types exhibits state-switching behaviour \citep{Tennessen:2019b}. On a fine scale, these profiles display bouts of short-term periodicity within each dive \citep{Adam:2019}. Fine-scale periodicity nested within a coarser state-switching process is also common in fields ranging from machine health \citep{Xin:2018,Lucero:2019} to speech recognition \citep{Juang:1991}.

@@

Our goal is to identify and describe discrete behavioural states at multiple scales within functional data (e.g., coarse-scale curve types and fine-scale behavioural states) and to model the dependence structure between those states. To accomplish this, we turn to the field of animal movement modelling \citep{Hooten:2017}, where one of the most prevalent techniques of late is the hidden Markov model, or HMM \citep{Patterson:2017,McClintock:2020}. HMMs interpret animal movement data as arising from a Markov chain of a discrete number of behavioural states, allowing biologists to infer the underlying behaviour of an animal from sequential observations of its position. While ubiquitous in ecology literature, HMMs have seen little use in non-parametric functional modelling with a few notable exceptions. In particular, \citet{Langrock:2018} take a non-parametric approach to model the distributions of HMM observations with B-splines, and \citet{DeSouza:2014,DeSouza:2017} use HMMs to model state-switching behaviour within functional data. However, none of these papers account for temporal correlation taking place on multiple scales.

>>>>

FDA was originally developed to process curves assumed to be independent replicates (i.e., there is no between-curve dependence), and within-curve fine-scale structure is not usually incorporated in FDA models. However, sets of curves often exhibit complex sequential dependencies both between curves and within curves, especially in the case of biologging data \citep{Barajas:2017}.
On a coarse scale, dive profiles of marine animals show a discrete number of distinct dive ``types"
%that aggregate together in time 
where the sequence of dive types exhibits state-switching behaviour \citep{Tennessen:2019b}. On a fine scale, these profiles can display bouts of short-term periodicity within each dive \citep{Adam:2019}. Fine-scale periodicity nested within a coarser state-switching process is also common in fields ranging from machine health \citep{Xin:2018,Lucero:2019} to speech recognition \citep{Juang:1991}.

@@

Our goal is to identify and describe discrete behavioural states at multiple scales within functional data (e.g., coarse-scale curve types and fine-scale behavioural states) and to model the dependence structure between those states. To accomplish this, we turn to the field of animal movement modelling \citep{Hooten:2017}, where one of the most prevalent techniques of late is the hidden Markov model, or HMM \citep{Patterson:2017,McClintock:2020}. HMMs interpret animal movement data as arising from a Markov chain of a discrete number of behavioural states, allowing biologists to infer the underlying behaviour of an animal from sequential observations of its position. While ubiquitous in ecology literature, HMMs have seen little use in non-parametric functional modelling with a few notable exceptions. In particular, \citet{Langrock:2018} take a non-parametric approach to model the distributions of HMM observations with B-splines, while \citet{DeSouza:2014} and \citet{DeSouza:2017} use HMMs to model state-switching behaviour within functional data. However, none of these papers account for temporal correlation taking place on multiple scales.

### 2021-07-09 Report/Sections/models.tex
Final edits

We take a frequentest approach in this paper, which involves obtaining maximum likelihood estimates $\{\hat \theta, \hat \Gamma\}$ from Equation (\ref{eqn:like_marginal}). The estimated probability of each hidden state $\hat{\Pr}(X_t = i|Y = y)$ can then be calculated using $\{\hat \theta, \hat \Gamma\}$ and the \textit{forward-backward algorithm} \citep{Zucchini:2016}. Alternatively, it is reasonable to use a Bayesian approach, where Equation (\ref{eqn:like}) is combined with a prior to obtain a posterior distribution over $X$, $\theta$, and $\Gamma$. This posterior distribution can be sampled from using methods such as sequential Monte Carlo \citep{Douc:2011b}, Markov-chain Monte Carlo \citep{Scott:2002}, or variational inference \citep{Foti:2014}.
%

@@

Replacing $Y^*_t$ with summary statistics results in a loss of information since it involves substituting infinite-dimensional curves with a finite-dimensional representation. While we are primarily interested in classifying and describing coarse- and fine-scale behaviour, some researchers may be less interested in interpretability and more interested in predicting future functional data. To this end, \citet{Aue:2015} and \citet{Gao:2019} both use functional principle component analysis (FPCA) and derive bounds on the error caused by dimension reduction. While these papers do not account for state-switching behaviour between curves or windows, deriving similar error bounds for FPCA within an HMM framework appears to be a promising direction for future research.

>>>>

We take a frequentest approach in this paper, which involves obtaining maximum likelihood estimates $\{\hat \theta, \hat \Gamma\}$ from Equation (\ref{eqn:like_marginal}). The estimated probability of each hidden state $\hat{\Pr}(X_t = i|Y = y)$ can then be calculated using $\{\hat \theta, \hat \Gamma\}$ and the \textit{forward-backward algorithm} \citep{Zucchini:2016}. Alternatively, it is reasonable to use a Bayesian approach, where Equation (\ref{eqn:like}) is combined with a prior distribution to obtain a posterior distribution over $X$, $\theta$, and $\Gamma$. This posterior distribution can be sampled from using methods such as sequential Monte Carlo \citep{Douc:2011b}, Markov-chain Monte Carlo \citep{Scott:2002}, or variational inference \citep{Foti:2014}.
%

@@

Replacing $Y^*_t$ with summary statistics results in a loss of information since it involves substituting infinite-dimensional curves with a finite-dimensional representation. While we are primarily interested in classifying and describing coarse- and fine-scale behaviour, some researchers may be less interested in interpretability and more interested in predicting future functional data. To this end, \citet{Aue:2015} and \citet{Gao:2019} both use functional principle component analysis (FPCA) and derive bounds on the reconstruction error caused by dimension reduction. While these papers do not account for state-switching behaviour between curves or windows, deriving similar error bounds for FPCA within an HMM framework appears to be a promising direction for future research.

### 2021-07-09 Report/Sections/results.tex
Final edits

The coarse-scale parameter estimates suggest that the killer whale has at least two distinct dive behaviours (see Table \ref{table:emis_dists_CarHHMM-DFT} and Figure \ref{fig:emis}). 
Dive type 1 corresponds to shorter, shallower dives which likely reflect resting, travelling, and to a lesser extent searching for prey.
Dive type 2 is longer, deeper, and may be associated with behaviours such as hunting  \citep{Tennessen:2019b}, but it is unclear whether any of the dives in this study are successful foraging dives. No dive in this data set has a maximum depth deeper than 30 meters and a study by \citet{Wright:2017} of killer whales in Johnstone Strait found that most prey captures occur at depths deeper than 100 meters. However, the killer whale studied here was tagged north of Johnstone Strait in Queen Charlotte Sound, and %researchers 
unpublished data collected by the authors suggest that %significant numbers of 
prey-capture events can occur near the surface. %\citep{Fortune:2020}. 
Dive type 2 could also be associated with behaviours such as socializing that can take place several meters below the surface \citep{Tennessen:2019b}.

@@

We use two visual tools to evaluate the CarHHMM-DFT: pseudoresidual plots and empirical histograms. The pseudoresidual of a coarse-scale observation $y_t$ is equal to $\Phi^{-1} \left(Pr(Y_t < y_t|\{Y_1,\ldots,Y_T,\Z_1,\ldots,\Z_T\}/\{Y_t\}) \right)$ and the pseudoresidual of a fine-scale observation $\z_{t,\tilde t^*}$ is $\Phi^{-1} \left(Pr(\Z_{t,t^*} < \z_{t,\tilde t^*}|\{Y_1,\ldots,Y_T,\Z_1,\ldots,\Z_T\}/\{Y^*_{t,\tilde t}\}) \right)$, where $\Phi$ is the cumulative distribution function of a standard Normal distribution. If the model is correct, then all pseudoresiduals are independent and follow a standard Normal distribution. Histograms of the pseudoresiduals mostly support that the CarHHMM-DFT is well-specified. One exception is $\Ztwo_{t,\tilde t^*}$, whose pseudoresiduals are noticeably right-skewed (see Figure \ref{fig:pseudoresids}). This implies that the true distribution of $\Ztwo_{t,\tilde t^*}$ may follow a heavier-tailed distribution than the Gamma distribution used in the case study. See Sections 5 through 7 of supplement A for pseudoresidual plots corresponding to all observations and models.

We also plot histograms of dive duration corresponding to each dive type in Figure \ref{fig:empirical_dist}. Each observation of dive duration is weighted by the estimated probability that it corresponds to a particular dive type as decoded by the %\textit{forward-backward algorithm} \citep{Zucchini:2016}. 
forward-backward algorithm. This procedure results in two histograms -- one corresponding to dive type 1 and another corresponding to dive type 2. Each histogram is plotted together with the corresponding emission distribution estimated by the CarHHMM-DFT. Analogous histograms corresponding to the fine-scale observations are contained in Section 6 and 7 of of supplement A. Our results mostly show that the CarHHMM-DFT explains the data well, but there are some exceptions. In particular, histograms corresponding to subdive state 3 show that $\Zone_{t,\tilde t^*}$ has heavier tails compared to a normal distribution. This indicates the existence of rare events corresponding to exceptionally sudden changes in acceleration of the killer whale. These outliers are potential subjects for future study and may indicate biologically relevant phenomena such as prey capture \citep{Tennessen:2019a}.

>>>>

The coarse-scale parameter estimates suggest that the killer whale has at least two distinct dive behaviours (see Table \ref{table:emis_dists_CarHHMM-DFT} and Figure \ref{fig:emis}). 
Dive type 1 corresponds to shorter, shallower dives which likely reflect resting, travelling, and to a lesser extent searching for prey.
Dive type 2 is longer, deeper, and may be associated with behaviours such as hunting  \citep{Tennessen:2019b}, but it is unclear whether any of the dives in this study are successful foraging dives. No dive in this data set has a maximum depth deeper than 30 meters, and a study by \citet{Wright:2017} of killer whales in Johnstone Strait found that most prey captures occur at depths deeper than 100 meters. However, the killer whale studied here was tagged north of Johnstone Strait in Queen Charlotte Sound, and %researchers 
unpublished data collected by the authors suggest that %significant numbers of 
prey-capture events can occur near the surface. %\citep{Fortune:2020}. 
Dive type 2 could also be associated with behaviours such as socializing that can take place several meters below the surface \citep{Tennessen:2019b}.

@@

We use two visual tools to evaluate the CarHHMM-DFT: pseudoresidual plots and empirical histograms. The pseudoresidual of a coarse-scale observation $y_t$ is equal to $\Phi^{-1} \left(Pr(Y_t < y_t|\{Y_1,\ldots,Y_T,\Z_1,\ldots,\Z_T\}/\{Y_t\}) \right)$ and the pseudoresidual of a fine-scale observation $\z_{t,\tilde t^*}$ is $\Phi^{-1} \left(Pr(\Z_{t,t^*} < \z_{t,\tilde t^*}|\{Y_1,\ldots,Y_T,\Z_1,\ldots,\Z_T\}/\{Y^*_{t,\tilde t}\}) \right)$, where $\Phi$ is the cumulative distribution function of a standard Normal distribution. If the model is correct, then all pseudoresiduals are independent and follow a standard Normal distribution. Histograms of the pseudoresiduals mostly support that the CarHHMM-DFT is well-specified. One exception is $\Ztwo_{t,\tilde t^*}$, whose pseudoresiduals are noticeably right-skewed (see Figure \ref{fig:model_checking}). This implies that the true distribution of $\Ztwo_{t,\tilde t^*}$ may follow a heavier-tailed distribution than the Gamma distribution used in the case study. See Sections 5 through 7 of supplement A for pseudoresidual plots corresponding to all observations and models.

We also plot histograms of dive duration corresponding to each dive type in Figure \ref{fig:model_checking}. Each observation of dive duration is weighted by the estimated probability that it corresponds to a particular dive type as decoded by the %\textit{forward-backward algorithm} \citep{Zucchini:2016}. 
forward-backward algorithm. This procedure results in two histograms -- one corresponding to dive type 1 and another corresponding to dive type 2. Each histogram is plotted together with the corresponding emission distribution estimated by the CarHHMM-DFT. Analogous histograms corresponding to the fine-scale observations are contained in Section 6 and 7 of of supplement A. Our results mostly show that the CarHHMM-DFT explains the data well, but there are some exceptions. In particular, histograms corresponding to subdive state 3 show that $\Zone_{t,\tilde t^*}$ has heavier tails compared to a normal distribution. This indicates the existence of rare events corresponding to exceptionally sudden changes in acceleration of the killer whale. These outliers are potential subjects for future study and may indicate biologically relevant phenomena such as prey capture \citep{Tennessen:2019a}.

### 2021-07-09 Report/Sections/simulation.tex
Final edits

The two simulated dive types differ in that dives of type 1 are much shorter on average (27 seconds) than dives of type 2 (128 seconds). The three simulated subdive states differ primarily because $\mu_W^*$ and $\sigma_W^*$ are much higher for subdive state 3 than for subdive state 2, which in turn is much higher than for subdive state 1. These larger parameter values correspond to much more vigorous and variable periodic behaviour in the acceleration data.

>>>>

The two simulated dive types differ in that dives of type 1 are much shorter on average (27 seconds) than dives of type 2 (128 seconds). The three simulated subdive states differ primarily because $\mu_W^*$ and $\sigma_W^*$ are much higher for subdive state 3 than for subdive state 2, which in turn are much higher than for subdive state 1. These larger parameter values correspond to much more vigorous and variable periodic behaviour in the acceleration data.
